Mega-analysis of the Interoceptive Accuracy Scale (IAS) Structure and its Subjective Correlates

### 2025-06-21 analysis/Study1.qmd
edits to paper

---
title: "Meta-analysis of the Interoceptive Accuracy Scale (IAS)"
subtitle: "Study 1"
editor: source
editor_options: 
  chunk_output_type: console
format:
  html:
    code-fold: true
    self-contained: true
    toc: true
    fig-dpi: 100
    fig-width: 11.33
execute: 
  cache: true
---

@@

# Create the dataframe
table <- data.frame(
   Sample = c('Murphy et al., (2020)', '', '', 'Gaggero et al., (2021)', 'Campos et al., (2022)', 'Todd et al., (2022)', 'Arslanova et al., (2022)', 'Brand et al., (2022)', 'Brand et al., (2023)', '', '', '', 'Lin et al., (2023)', '', '', 'VonMohr et al., (2023)', 'Makowski et al., (2023a)', 'Makowski et al., (2023b)', 'Makowski et al., (2023c)', 'Poreiro et al., (2024)', 'Poreiro et al., unpublished', 'Total'),
  Subsample = c('', 'Sample 1', 'Sample 2', '', '', '', '', '', '', 'Sample 1', 'Sample 2', 'Sample 3', '', 'Sample 1', 'Sample 2', '', '', '', '', '','', ''),   
  Language = c('', 'English', 'English' , 'English and Italian', 'Portuguese', 'English', 'English', 'German', '', 'German', 'German', 'German', '', 'Chinese', 'Chinese', 'English', 'English', 'English', 'English', 'English', 'English',''),
  N = c('', nrow(df1a), nrow(df1b), nrow(df2), nrow(df3), nrow(df4), nrow(df5_attention), nrow(df6),'', nrow(df7a), nrow(df7b), 802,'', nrow(df8a), nrow(df8b), nrow(df9), nrow(df10a), nrow(df10b), nrow(df10c), nrow(df11a), nrow(df11b), 32214),
  'Difference' = c('','', '', '', '', '', '', '', '', '', '', '', '', 'Collapsed "Itch" and "Tingling"', 'Collapsed "Itch" and "Tingling"', '', 'Analog scales', 'Analog scales', 'Analog scales', '', '', ''),
  Age = c('', make_age(df1a$Age), make_age(df1b$Age), make_age(df2$Age), make_age(df3$Age), "48.6.6 ± 14.1*", make_age(df5_attention$Age), make_age(df6$Age),'', make_age(df7a$Age), make_age(df7b$Age), make_age(df7c$Age),'', make_age(df8a$Age), make_age(df8b$Age), make_age(df9$Age), make_age(df10a$Age), make_age(df10b$Age), make_age(df10c$Age), make_age(df11a$Age), make_age(df11b$Age), '48.6 ± 13.1'),
  Range = c('', '18-69', '18-91', '18-58', '18-72', '18-92*', '18-73', '18-78', '', '18-79', '16-81', '18-72','', '16-60', '20-60', '18-93', '18-73', '17-76','18-50', '18-57', '18-60', '17-93'),
  Female_Percentage = c('', '69.4%', '70.1%', '60.3%', '59.6%', '50%*', '46.8%', '78.7%', '', '79.5%', '77.7%', '68.9%', '', '57.0%', '56.2%', '73.2%', '50.3%', '53.0%', '76%', '74.8%', '75.9%', '71.6%'),
  Availability= c('osf.io/3m5nh', '', '', 'osf.io/5x9sg', 'osf.io/j6ef3', 'osf.io/ms354', 'osf.io/mp3cy', 'osf.io/xwz6g', 'osf.io/3f2h6', '', '', '','osf.io/3eztd','', '', 'osf.io/7p9u5', 'github.com/RealityBending/IllusionGameReliability', 'github.com/DominiqueMakowski/PHQ4R', 'github.com/RealityBending/InteroceptionPrimals', 'osf.io/49wbv', '','')
)

@@

@@

p1 <- p1_data |>
  mutate(Sample = fct_relevel(Sample, "Sample 1a", "Sample 1b", "Sample 2", "Sample 3", "Sample 4", "Sample 5", "Sample 6", "Sample 7a", "Sample 7b", "Sample 7c", "Sample 8a", "Sample 8b", "Sample 9", "Sample 10a", "Sample 10b", "Sample 10c", "Sample 11a", "Sample 11b")) |>
  # mutate(Dashed = ifelse(Parameter %in% c("Affective_touch", "Blood_Sugar", "Bruise"), TRUE, FALSE)) |> 
  ggplot(aes(x = x, y = y, color = Parameter)) +
  geom_line() +
  scale_color_material_d() +
  scale_linetype_manual(values = c("solid", "dashed"), guide="none") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        plot.title = element_text(face="bold")) +
  labs(x = "Response", y = "Density", color = "Item") +
  facet_wrap(~Sample, scales = "free_y", nrow = 5)

ggsave("figures/Figure1.png", p1, width=10, height=10, dpi=200, bg="white")
p1

>>>>

---
title: "Mega-analysis of the Interoceptive Accuracy Scale (IAS)"
subtitle: "Study 1"
editor: source
editor_options: 
  chunk_output_type: console
format:
  html:
    code-fold: true
    self-contained: true
    toc: true
    fig-dpi: 100
    fig-width: 11.33
execute: 
  cache: true
---

@@

# Create the dataframe
table <- data.frame(
   Sample = c('Murphy et al., (2020)', '', '', 'Gaggero et al., (2021)', 'Campos et al., (2022)', 'Todd et al., (2022)', 'Arslanova et al., (2022)', 'Brand et al., (2022)', 'Brand et al., (2023)', '', '', '', 'Lin et al., (2023)', '', '', 'VonMohr et al., (2023)', 'Makowski et al., (2023a)', 'Makowski et al., (2023b)', 'Makowski et al., (2023c)', 'Poerio et al., (2024)', 'Poerio et al., unpublished', 'Total'),
  Subsample = c('', 'Sample 1', 'Sample 2', '', '', '', '', '', '', 'Sample 1', 'Sample 2', 'Sample 3', '', 'Sample 1', 'Sample 2', '', '', '', '', '','', ''),   
  Language = c('', 'English', 'English' , 'English and Italian', 'Portuguese', 'English', 'English', 'German', '', 'German', 'German', 'German', '', 'Chinese', 'Chinese', 'English', 'English', 'English', 'English', 'English', 'English',''),
  N = c('', nrow(df1a), nrow(df1b), nrow(df2), nrow(df3), nrow(df4), nrow(df5_attention), nrow(df6),'', nrow(df7a), nrow(df7b), 802,'', nrow(df8a), nrow(df8b), nrow(df9), nrow(df10a), nrow(df10b), nrow(df10c), nrow(df11a), nrow(df11b), 32214),
  'Difference' = c('','', '', '', '', '', '', '', '', '', '', '', '', 'Collapsed "Itch" and "Tingling"', 'Collapsed "Itch" and "Tingling"', '', 'Analog scales', 'Analog scales', 'Analog scales', '', '', ''),
  Age = c('', make_age(df1a$Age), make_age(df1b$Age), make_age(df2$Age), make_age(df3$Age), "48.6.6 ± 14.1*", make_age(df5_attention$Age), make_age(df6$Age),'', make_age(df7a$Age), make_age(df7b$Age), make_age(df7c$Age),'', make_age(df8a$Age), make_age(df8b$Age), make_age(df9$Age), make_age(df10a$Age), make_age(df10b$Age), make_age(df10c$Age), make_age(df11a$Age), make_age(df11b$Age), '48.6 ± 13.1'),
  Range = c('', '18-69', '18-91', '18-58', '18-72', '18-92*', '18-73', '18-78', '', '18-79', '16-81', '18-72','', '16-60', '20-60', '18-93', '18-73', '17-76','18-50', '18-57', '18-60', '17-93'),
  Female_Percentage = c('', '69.4%', '70.1%', '60.3%', '59.6%', '50%*', '46.8%', '78.7%', '', '79.5%', '77.7%', '68.9%', '', '57.0%', '56.2%', '73.2%', '50.3%', '53.0%', '76%', '74.8%', '75.9%', '71.6%'),
  Availability= c('osf.io/3m5nh', '', '', 'osf.io/5x9sg', 'osf.io/j6ef3', 'osf.io/ms354', 'osf.io/mp3cy', 'osf.io/xwz6g', 'osf.io/3f2h6', '', '', '','osf.io/3eztd','', '', 'osf.io/7p9u5', 'github.com/RealityBending/IllusionGameReliability', 'github.com/DominiqueMakowski/PHQ4R', 'github.com/RealityBending/InteroceptionPrimals', 'osf.io/49wbv', '','')
)

@@

linetype <- setNames(rep("solid", length(vars)), vars)
linetype["Affective touch"] <- "dotted"
linetype["Blood sugar"] <- "dashed"
linetype["Sex arousal"] <- "solid"
linetype["Bruise"] <- "dashed"

@@

p1 <- p1_data |>
  mutate(Sample = fct_relevel(Sample, "Sample 1a", "Sample 1b", "Sample 2", "Sample 3", "Sample 4", "Sample 5", "Sample 6", "Sample 7a", "Sample 7b", "Sample 7c", "Sample 8a", "Sample 8b", "Sample 9", "Sample 10a", "Sample 10b", "Sample 10c", "Sample 11a", "Sample 11b"),
         Parameter = ifelse(Parameter == "Affective_touch", "Affective touch", Parameter),
         Parameter = ifelse(Parameter == "Blood_Sugar", "Blood sugar", Parameter),
         Parameter = ifelse(Parameter == "Sex_arousal", "Sex arousal", Parameter)) |>
  # mutate(Dashed = ifelse(Parameter %in% c("Affective_touch", "Blood_Sugar", "Bruise"), TRUE, FALSE)) |> 
  ggplot(aes(x = x, y = y, color = Parameter)) +
  geom_line(aes(linetype = Parameter), linewidth = 0.7) +
  scale_color_material_d() +
  scale_linetype_manual(values = linetype) +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        plot.title = element_text(face="bold")) +
  labs(x = "Response", y = "Distribution", color = "Item", linetype = "Item") +
  facet_wrap(~Sample, scales = "free_y", nrow = 3)

ggsave("figures/Figure1.png", p1, width=12, height=7, dpi=300, bg="white")
p1
```

### 2025-06-21 paper/1_Initial/manuscript.qmd
edits to paper

---
title: "Mega-analysis of the Interoceptive Accuracy Scale (IAS) Structure and its Subjective Correlates"
shorttitle: "IAS Meta-analysis"
author:
  - name: Ana Neves
    orcid: 0009-0006-0020-7599
    roles:
      - Data curation
      - Formal Analysis
      - Investigation
      - Visualization
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - id: "id1"
        name: "University of Sussex"
        department: "School of Psychology"
        city: "Brighton"
        country: "UK"
        postal-code: "BN1 9RH"
# author-note:
#   status-changes: 
#     affiliation-change: null
#     deceased: null
#   disclosures:
#     study-registration: null
#     data-sharing: null
#     related-report: null
#     conflict-of-interest: null
#     financial-support: null
#     gratitude: null
#     authorship-agreements: null
    #   # - Project administration
    #   # - Data curation
    #   # - Formal Analysis
    #   # - Investigation
    #   # - Visualization
    #   # - Writing – original draft
    #   # - Writing – review & editing
    # Roles are optional. 
    # conceptualization, data curation, formal Analysis, funding acquisition, investigation, 
    # methodology, project administration, resources, software, supervision, validation, 
    # visualization, writing, editing
    #   - Conceptualization
    #   - Data curation
    #   - formal Analysis
    #   - Funding acquisition
    #   - Investigation
    #   - Methodology
    #   - Project administration
    #   - Resources
    #   - Software
    #   - Supervision
    #   - Validation
    #   - Visualization
    #   - Writing – original draft
  - name: Magdalena Pfaff
    orcid: 0009-0006-2386-7936
    roles:
      - Data curation
      - Writing – original draft
    affiliations:
      - ref: "id1"
  - name: Dominique Makowski
    corresponding: true
    orcid: 0000-0001-5375-9967
    email: D.Makowski@sussex.ac.uk
    url: https://realitybending.github.io/
    roles:
      - Project administration
      - Data curation
      - Formal Analysis
      - Investigation
      - Visualization
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - ref: "id1"
      - name: "University of Sussex"
        department:  "Sussex Centre for Consciousness Science"
author-note:
  disclosures:
    gratitude: |
      ::: {.callout-note icon=false appearance="simple"}
      This preprint is a non-peer-reviewed work from the [**Reality Bending Lab**](https://realitybending.github.io/).
      ![](https://realitybending.github.io/media/ReBeL_LogoOnly_hu11484441381606756729.png){width=20% fig-align="center"}
      :::
abstract: |
  Blabla the abstract blabla.
keywords: [keyword1, keyword2, keyword3]
floatsintext: true
numbered-lines: true
bibliography: bibliography.bib
suppress-title-page: false
mask: false
# Language options. See https://quarto.org/docs/authoring/language.html
lang: en
language:
  citation-last-author-separator: "and"
  citation-masked-author: "Masked Citation"
  citation-masked-date: "n.d."
  citation-masked-title: "Masked Title"
  title-block-author-note: "Author Note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
  references-meta-analysis:
format:
  apaquarto-pdf:
    # Can be jou (journal), man (manuscript), stu (student), or doc (document)
    documentmode: jou
    include-in-header:
      - text: |
          \usepackage{lscape}
  apaquarto-docx: default
  apaquarto-html: default
editor: 
  markdown: 
    wrap: sentence
editor_options: 
  chunk_output_type: console
---

## Introduction

@@

### The Interoceptive Assessment Puzzle

@@

### The Interoceptive Accuracy Scale (IAS)

@@

## Study 1

Study 1 will re-analyse and assess the factor structure of the IAS by taking advantage of the large number of open-access datasets [@brand2023; @campos2021; @lin2023; @murphy2019; @arslanova2022; @todd2022; @gaggero2021; @brand2022; @von2023].
While combining these studies might provide a more robust and generalizable understanding of the IAS' factor structure, we also additionally provide an individual analysis (i.e., on all samples separately) to add nuance to the general picture, as all studies differ in their sample size, demographic characteristics, language, and procedure.

### Methods

#### Datasets

Our search focused on studies citing the original IAS validation paper [@murphy2019], identifying 136 papers (as of 01/05/2024).
To qualify for inclusion, papers needed to (1) provide accessible data in open-access, (2) employ the IAS as a measure, and (3) report individual IAS items scores.
We also included the data of four unpublished (but already open-access) studies.
A total of 14 studies was included (see **Table 1**).

@@

#### Data Analysis

Psychometrically good items should exhibit various qualities, such as validity and reliability, and one of the contributing factors is the amount of variability captured by an item. Items to which all participants' answers are concentrated around one option - i.e., exhibiting a narrow distribution - should be flagged as potentially problematic.

After examining the distributions to all IAS items, we will analyze the factor structure using two different approaches, namely traditional exploratory and confirmatory Factor Analysis (EFA/CFA) as well as Exploratory Graph Analysis (EGA).

<!-- EGA -->
By combining network analysis with psychometric methods, the recently-developed EGA framework allows to jointly estimate the number of dimensions (i.e., groups of items), the structure as well as its stability [@golino2017exploratory; @golino2020investigating]. 
Evidence has underlined its suitability as an alternative to traditional factor analysis, addressing some of its limitations such as the assumption of a "latent" source of variability, possible biasing in the estimation of the optimal factor numbers depending on sample size, and the poor performance of other methods in complex population structures, while remaining comparable and interpretable [@christensen2021equivalency; @jimenez2023dimensionality]. 
At a fundamental level, EGA conceptualizes variables as nodes in a network, with connections (edges) reflecting associations between them. 
Clustering these nodes reveals distinct communities of related items, in practice akin to traditional latent factors - but without explicitly assuming their presence [@christensen2021equivalency].
<!-- By applying UVA, the analysis can maintain the integrity of the IAS while improving its efficiency and reducing redundancy. -->

<!-- UVA -->
To assess whether the IAS contains redundant items (e.g., due to multicollinearity or local dependency), Unique Variable Analysis [UVA, @christensen2023unique] will be employed. 
UVA is another novel network psychometric method designed to identify and merge items that share substantial variance, effectively reducing datasets to a set of unique variables. 
Unlike other reduction methods that aim to minimize variables broadly, UVA targets redundancy specifically, preserving meaningful information that may otherwise be lost [@merritt2022experimental]. 
By focusing on statistically redundant variables, UVA offers a middle ground between no reduction and extensive reduction approaches. 
Simulation studies have demonstrated that using a threshold of 0.25 optimally balances accuracy and false positives, ensuring that only genuinely redundant items are combined.

<!-- Factor Analysis -->

While EGA offers a robust alternative to traditional factor analysis, factor analysis remains a widely used method for dimensionality assessment, hence it will be also be computed in this analysis. 
Unlike EGA, factor analysis assumes a latent source of variability — a common latent variable — underlies the observed set of manifest variables [@cosemans2022exploratory].
A critical step in factor analysis is determining the optimal number of factors, typically achieved by examining eigenvalue patterns and applying stopping rules, such as the Kaiser-Guttman criterion (eigenvalues greater than 1) or parallel analysis.
These methods aim to identify the most meaningful factor structure, though their performance can vary depending on sample size and data complexity [@christensen2023unique].

To determine the appropriate number of factors to retain during factor analysis, the n_factors function from the performance package will be used [@ludecke2021performance]. 
This function runs multiple established procedures for factor retention and identifies the optimal number of factors based on the maximum consensus across methods - i.e., Method Agreement Procedure. 
These methods include the Kaiser criterion, parallel analysis,  Velicer's MAP, and others.
Following this, EFA with the optimal number of factors identified, using 'oblimin' rotation, will be aplied. 
Model fit using CFA to identify the best-fitting model will then be computed.

### Results

@@

The distribution of the items across samples suggests the presence of a consistent modal value (@fig-distributions).
In other words, participants are most likely to answer 4/5 (i.e., agree) on most items, with the exception of "blood sugar" and "bruise", which exhibit a different distributional pattern with a lower mode (\~2/5). 
This pattern persists across most samples, except for "affective touch" in samples 8a and 8b, where it deviates from the general trend. 
In these samples, "affective touch" follows a distribution similar to that of "blood sugar" and "bruise".  **-could it be a language thing?**
Additionally, one can note the low occurrence of extreme values (1 and 5), meaning that the bulk of answers varies between 3 values (assuming the IAS is implemented as a 5-point Likert scale following its validation).
The samples using an analogue scale (samples 10a, 10b and 10c in the figure) displaying a more continuous and progressive spread of answers, seemingly improving the interindividual variability, although potentially suggesting a second lower mode at \~2.  **- sample 11's were rated on a liker-scale**

**TODO: And add x-axis on all subplots. Consider making the problematic items dashed.**.

```{r}
#| warning: false
#| label: fig-distributions
#| fig-cap: "Distribution of responses across datasets reveals a consistent modal value, typically around 4 or 5 (indicating agreement), except for 'blood sugar' and 'bruise,' which have lower modes. Most responses cluster around the middle values, with few extreme scores (1 and 5). Samples using an analogue scale (10a, 10b, 10c) show a more continuous distribution and increased interindividual variability. Since most samples use Likert scales (discrete), the density plots may not be the most accurate representation but were chosen to clearly highlight variability patterns in the data."
#|
knitr::include_graphics("figures/Figure1.png")
```

<!-- Univariate analysis -->
UVA flagged two strongly redundant variables, "itch" and "tickle" - suggesting to remove the latter.
Several more pairs of items were flagged as moderately redundant ("wind" and "burp"; "urinate" and "defecate") and mildly redundant ("sneeze" and "cough"; "heart" and "breathing"; "hungry" and "thirsty").
These patterns consistently appeared in most samples when considered individually.

<!-- Network analysis and Strcuture stability -->

**TO DOM: I put a bunch of info from the fucntion you created an the package documentation. Let me know what you want to keep from this information**

The EGA analysis employed a sparse Gaussian graphical model with the graphical lasso (glasso) method. 
To identify community structures, both the Louvain and Walktrap algorithms were applied.
The Walktrap algorithm identified an optimal solution with five clusters, whereas the Louvain algorithm indicated an optimal solution with four clusters. 
To assess the stability of these solutions, a bootstrap resampling approach was implemented, evaluating stability based on the proportion of bootstrap samples in which the original structure was replicated.
The four-cluster solution demonstrated greater item stability, with a higher proportion of items consistently appearing within the same dimension across bootstrap samples.
In this solution, item stability for all variables exceeded 0.91, with the exception of "affective touch" in the four-cluster solution, which showed comparatively lower stability.
Nevertheless, this solution was considered the most optimal; C1. "itch", "tickle", "bruise", "blood sugar"; C2. "burp", "wind", "cough", "sneeze", "vomit"; C3. "affective touch", "sexual arousal", "muscles", "temperature", "pain", and "taste"; C4. "heart", "breathing", "hungry", "thirsty", "urinate", and "defecate". 
This solution was generally consistent within the data sets, although some samples yielded 3- or 5- clusters as the optimal structure.

**EGA Clusters not consistent with FA dimensions**
<!-- EGA clusters 
c1- "itch", "tickle", "bruise", "blood sugar
c2- burp", "wind", "cough", "sneeze", "vomit"
c3 - affective touch", "sexual arousal", "muscles", "temperature", "pain", and "taste"
c4- "heart", "breathing", "hungry", "thirsty", "urinate", and "defecate"

@@

**TODO: explain how we obtained the optimal number of factors** -->  **I think i explained it above on the analysis bit, please confirm whether there needs to be more info**

Across all samples, the Method Agreement Procedure recommended a four-factor solution. 
Approximately 30% of the tested methods — the Beta method, Optimal Coordinates, Parallel Analysis, and the Kaiser criterion — converged on this as the optimal number of factors to retain.

In line with the Method Agreement Procedure's recommendation, a four-factor solution was extracted using an oblimin rotation, accounting for 41.58% of the total variance.
The first factor included items "burp," "cough," "wind," "sneeze," "vomit," "temperature," "sexual arousal," and "taste," explaining 14.45% of the total variance. 
The second factor incuded "breathing," "hungry," "heart," "thirsty," "pain," "muscles," "blood sugar," and "bruise," accounting for 11.76% of the variance. 
The third factor contained "tickle," "itch," and "affective touch," explaining 8.09% of the total variance. 
Lastly, the fourth factor included "urinate" and "defecate," accounting for only 7.28% of the total variance.
Notably this grouping differs from that of the EGA's four cluster optimal solution.

<!-- CFA -->
Throughout the EGA and EFA analyses, the item "tickle" was consistently identified as redundant and was flagged for removal in the UVA analysis. 
It also demonstrated the lowest uniqueness value in the factor analysis, leading to its exclusion from further analysis.
Similarly, several "ambiguous" items — such as "temp", "vomit", "affective touch", "sexual arousal", and "taste" — were removed due to their context-dependent nature as well as weak loadings on multiple factors, which may compromise measurement consistency and reduce the clarity of factor structure.

@@

### Discussion

@@

## Study 2

@@

### Methods

#### Materials

@@

#### Data Analysis

@@

### Results

#### Average correlations

@@

##### Correlations with body measures

@@

##### Correlations with mood measures

@@

##### Correlations with psychopathology measures

@@

##### Correlations with personality measures

@@

##### Correlations with other measures

@@

## General Discussion

@@

## Limitations and Future Directions

@@

## Conclusion

@@

## References

>>>>

---
title: "Mega-analysis of the Interoceptive Accuracy Scale (IAS) Structure and its Subjective Correlates"
shorttitle: "IAS Meta-analysis"
author:
  - name: Ana Neves
    orcid: 0009-0006-0020-7599
    roles:
      - Data curation
      - Formal Analysis
      - Investigation
      - Visualization
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - id: "id1"
        name: "University of Sussex"
        department: "School of Psychology"
        city: "Brighton"
        country: "UK"
        postal-code: "BN1 9RH"
# author-note:
#   status-changes: 
#     affiliation-change: null
#     deceased: null
#   disclosures:
#     study-registration: null
#     data-sharing: null
#     related-report: null
#     conflict-of-interest: null
#     financial-support: null
#     gratitude: null
#     authorship-agreements: null
    #   # - Project administration
    #   # - Data curation
    #   # - Formal Analysis
    #   # - Investigation
    #   # - Visualization
    #   # - Writing – original draft
    #   # - Writing – review & editing
    # Roles are optional. 
    # conceptualization, data curation, formal Analysis, funding acquisition, investigation, 
    # methodology, project administration, resources, software, supervision, validation, 
    # visualization, writing, editing
    #   - Conceptualization
    #   - Data curation
    #   - formal Analysis
    #   - Funding acquisition
    #   - Investigation
    #   - Methodology
    #   - Project administration
    #   - Resources
    #   - Software
    #   - Supervision
    #   - Validation
    #   - Visualization
    #   - Writing – original draft
  - name: Magdalena Pfaff
    orcid: 0009-0006-2386-7936
    roles:
      - Data curation
      - Writing – original draft
    affiliations:
      - ref: "id1"
  - name: Dominique Makowski
    corresponding: true
    orcid: 0000-0001-5375-9967
    email: D.Makowski@sussex.ac.uk
    url: https://realitybending.github.io/
    roles:
      - Project administration
      - Data curation
      - Formal Analysis
      - Investigation
      - Visualization
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - ref: "id1"
      - name: "University of Sussex"
        department:  "Sussex Centre for Consciousness Science"
author-note:
  disclosures:
    gratitude: |
      ::: {.callout-note icon=false appearance="simple"}
      :::
      This preprint is a non-peer-reviewed work from the [**Reality Bending Lab**](https://realitybending.github.io/).
      ![](https://realitybending.github.io/media/ReBeL_LogoOnly_hu11484441381606756729.png){width=20% fig-align="center"}
abstract: |
  Blabla the abstract blabla.
keywords: [keyword1, keyword2, keyword3]
floatsintext: true
numbered-lines: true
bibliography: bibliography.bib
suppress-title-page: false
mask: false
# Language options. See https://quarto.org/docs/authoring/language.html
lang: en
language:
  citation-last-author-separator: "and"
  citation-masked-author: "Masked Citation"
  citation-masked-date: "n.d."
  citation-masked-title: "Masked Title"
  title-block-author-note: "Author Note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
  references-meta-analysis:
format:
  apaquarto-pdf:
    # Can be jou (journal), man (manuscript), stu (student), or doc (document)
    documentmode: jou
    include-in-header:
      - text: |
          \usepackage{lscape}
  apaquarto-docx: default
  apaquarto-html: default
editor: 
  markdown: 
    wrap: sentence
editor_options: 
  chunk_output_type: console
---

# Introduction

@@

## The Interoceptive Assessment Puzzle

@@

## The Interoceptive Accuracy Scale (IAS)

@@

# Study 1

Study 1 will re-analyse and assess the factor structure of the IAS by taking advantage of the large number of open-access datasets [@brand2023; @campos2021; @lin2023; @murphy2019; @arslanova2022; @todd2022; @gaggero2021; @brand2022; @von2023].
While combining these studies might provide a more robust and generalizable understanding of the IAS' factor structure, we also additionally provide an individual analysis (i.e., applying the same method on all samples separately) to add nuance to the general picture, as all studies differ in their sample size, demographic characteristics, language, and procedure.

## Methods

### Datasets

Our search focused on studies citing the original IAS validation paper [@murphy2019], identifying 136 papers (as of 01/05/2024).
To qualify for inclusion, papers needed to (1) provide accessible data in open-access, (2) employ the IAS as a measure, and (3) report individual IAS items scores.
We also included the data of four unpublished studies.
A total of 14 datasets was included (see **Table 1**).

@@

### Data Analysis

Psychometrically good items should exhibit various qualities, such as validity, reliability and discrimination, to which one of the contributing factors is the amount of inter-individual variability captured by an item. Items to which all participants' answers are concentrated around one option - i.e., exhibiting a narrow distribution - will be flagged as potentially problematic.

<!-- UVA -->
After examining the distributions to all IAS items, we will test for "redundant" items (e.g., due to multicollinearity or local dependency) using Unique Variable Analysis [UVA, @christensen2023unique], a novel method derived from network psychometric designed to identify and merge items that share substantial variance (which can distort the structure estimation).
We will use a threshold of 0.25, suggested to optimally balances accuracy and false positives **[REF NEEDED]**.

<!-- EGA -->
Following the analysis of items, we will analyze the factor structure of the IAS using two different approaches, namely traditional exploratory and confirmatory Factor Analysis (EFA/CFA) and Exploratory Graph Analysis (EGA).
By combining network analysis with psychometric methods, the recently-developed EGA framework allows to jointly estimate the number of dimensions (i.e., groups of items), the structure as well as its stability [@golino2017exploratory; @golino2020investigating]. 
Evidence has underlined its suitability as an alternative to traditional factor analysis, addressing some of its limitations such as the assumption of a "latent" source of variability, possible biasing in the estimation of the optimal factor numbers depending on sample size, and the poor performance of other methods in complex population structures, while remaining comparable and interpretable [@christensen2021equivalency; @jimenez2023dimensionality]. 
At a fundamental level, EGA conceptualizes variables as nodes in a network, with connections (edges) reflecting associations between them. 
Clustering these nodes reveals distinct communities of related items, in practice akin to traditional latent factors - but without explicitly assuming their presence [@christensen2021equivalency].

<!-- Factor Analysis -->
While EGA offers a robust alternative to traditional factor analysis, factor analysis remains a widely used method for dimensionality assessment. As our goal is to provide a general - yet nuanced - picture, with room to show potential discrepancies emerging from the methods used, we will also include it in the present study.
Unlike EGA, factor analysis assumes that  a latent source of variability — a common latent variable — underlies the observed set of manifest variables [@cosemans2022exploratory].
A critical step in factor analysis is determining the optimal number of factors, for which we will used the Method Agreement Procedure [@ludecke2021performance], which involves a consensus-based decision based on multiple factor estimation methods applied concurrently.

## Results

@@

The distribution of the items across samples suggests the presence of a consistent modal value (@fig-distributions).
In other words, participants are most likely to answer 4/5 (i.e., agree) on most items, with the exception of "blood sugar" and "bruise", which exhibit a different distributional pattern with a lower mode (\~2/5). 
This is also the case for "affective touch" in samples 8a and 8b (the Chinese validation samples), which might indicate localisation issues.
Additionally, one can note the low occurrence of extreme values (1 and 5), meaning that the bulk of answers varies between 3 values (assuming the IAS is implemented as a 5-point Likert scale following its validation).
The samples using an analogue scale (samples 10a, 10b and 10c in the figure) display a more continuous and progressive spread of answers, seemingly improving the interindividual variability, although potentially displaying a secondary lower mode at \~2 (which might be suggesting the existence of potential clusters of participants).

```{r}
#| warning: false
#| label: fig-distributions
#| apa-twocolumn: true  # A Figure Spanning Two Columns When in Journal Mode
#| out-width: "100%"
#| fig-cap: "Distribution of responses across datasets reveals a consistent modal value, typically around 4 or 5 (indicating agreement), except for 'blood sugar' and 'bruise' - in dashed lines - and 'affective touch' (dotted lines) in the Chinese validation sample, which have lower modes. Most responses cluster around the middle values, with few extreme scores (1 and 5). Samples using an analogue scale (10a, 10b, 10c) show a more continuous distribution and increased interindividual variability. Since most samples use Likert scales (discrete), the density plots may not be the most accurate representation but were chosen to clearly highlight variability patterns in the data."

knitr::include_graphics("../../analysis/figures/Figure1.png")
```

<!-- UVA -->
UVA flagged two strongly redundant variables, "itch" and "tickle" - suggesting to remove the latter.
Several more pairs of items were flagged as moderately redundant ("wind" and "burp"; "urinate" and "defecate") and mildly redundant ("sneeze" and "cough"; "heart" and "breathing"; "hungry" and "thirsty").
These patterns consistently appeared in most samples when considered individually.

<!-- Network analysis and Structure stability -->
The EGA analysis employed a sparse Gaussian graphical model (GGM) with the graphical lasso (glasso) method. We applied the Walktrap (**REF**) and the Louvain (**REF**) algorithms for clusters identification, which yielded a five and four clusters solution, respectively. The four-cluster solution appeared as being more stable (based on the bootstrapped stability analysis: it had a higher proportion of items consistently assigned to the same cluster across bootstrap samples).
In this solution, item stability for all variables exceeded 0.90, except for "affective touch". The first cluster included skin-related items "itch", "tickle", "bruise", and "blood sugar"; the second cluster contained expulsion-related items "burp", "wind", "cough", "sneeze", and "vomit"; the third cluster (**NAME or interpretation?**) comprised "affective touch", "sexual arousal", "muscles", "temperature", "pain", and "taste"; and the fourth cluster included visceroceptive items like "heart", "breathing", "hungry", "thirsty", "urinate", and "defecate".
This solution was generally consistent when applied to samples separately.

Regarding Factor Analysis, the Method Agreement Procedure recommended a four-factor solution for the whole sample. A four-factor solution was extracted using an *oblimin* rotation, accounting for 41.58% of the total variance.
The first factor (14.45%) included items "burp", "cough", "wind", "sneeze", "vomit", "temperature", "sexual arousal" and "taste". The second factor (11.76%) included "breathing", "hungry", "heart", "thirsty", "pain", "muscles", "blood sugar", and "bruise". 
The third factor (8.09%) contained skin-related items "tickle", "itch" and "affective touch", and the final factor (7.28%) included "urinate" and "defecate".
Notably, this structure differed from that suggested by EGA.

<!-- EGA clusters 
c1- "itch", "tickle", "bruise", "blood sugar
c2- burp", "wind", "cough", "sneeze", "vomit"
c3 - affective touch", "sexual arousal", "muscles", "temperature", "pain", and "taste"
c4- "heart", "breathing", "hungry", "thirsty", "urinate", and "defecate"

@@

<!-- Item selection -->
**UVA**, then re-run EGA and EFA. Then CFA.

<!-- CFA -->
We then attempted at testing various candidate structures using confirmatory factor analysis (CFA).

Throughout the EGA and EFA analyses, the item "tickle" was consistently identified as redundant and was flagged for removal in the UVA analysis. 
It also demonstrated the lowest uniqueness value in the factor analysis, leading to its exclusion from further analysis.
Similarly, several "ambiguous" items — such as "temp", "vomit", "affective touch", "sexual arousal", and "taste" — were removed due to their context-dependent nature as well as weak loadings on multiple factors, which may compromise measurement consistency and reduce the clarity of factor structure.

@@

## Discussion

@@

# Study 2

@@

## Methods

### Materials

@@

### Data Analysis

@@

## Results

### Average correlations

@@

#### Correlations with body measures

@@

#### Correlations with mood measures

@@

#### Correlations with psychopathology measures

@@

#### Correlations with personality measures

@@

#### Correlations with other measures

@@

# General Discussion

@@

# Limitations and Future Directions

@@

# Conclusion

@@

# References
